Clouds and Radiation in a mock-Walker Circulation

### 2019-11-08 Silvers_WalkerCell.tex
changes to the conclusions and the section on resolution dependence

In contrast to the P100, P25, and E25 experiments just discussed, turning the LW CRE off for 
the E1 and E2 experiments results in more precipitation.  This can be understood as follows.  
The primary method by which the longwave cloud radiative effect influences the atmosphere is by 
heating the atmosphere in the region between the clouds and the surface.  
Larger values of ice condensate and upper-level cloud fraction as seen in the E2 and E1 experiment 
(Figure \ref{fig:TotCond}) therefore imply a larger atmospheric heating due to the CRE relative to the 
E25, P25, and P100 experiments in which have fewer clouds aloft (Figures \ref{fig:TotCond}, 
\ref{fig:TotCond_P25P100}, and \ref{fig:cf_tdtlw}).  When the warming effect of the upper level clouds 
in the E1 and E2 experiments is removed in the LWCRE off experiments the energy balance of 
the atmosphere must be maintained through an increase of latent heating and subsequent increase 
of precipitation.  This is reflected in Table 2.

@@

Water vapor is an efficient emitter of longwave radiation and thus greatly facilitates the cooling of the troposphere.  
\citeA{Popp2017} showed that there is dramatically less condensate in the atmosphere and much less precipitation 
(at the equator) for LW CRE off experiments (see their figure 1a,b).  Our results here for the E25, P25, and P100 
are consistent with those of \citeA{Popp2017} because the low-level condensate are strongly decreased 
when the LWCRE is off.  
%This is not the case for E2 and E1.  Those experiments have little low-level condensate to begin with 
%%and thus do not experience 
%%large changes in low-level condensate when the LWCRE is off.  
%but they do have a large increase of upper
%level condensate for the LWCRE off experiments.   
%This is a large difference in the way that models with explicit convection 
%respond to the LWCRE compared to more traditional large-scale models with parameterized convection.

The upper panels of Figures \ref{fig:rh_psi_P25E2E1} and \ref{fig:rh_psi_P25E2E1_lwoff} show that the 
interactions between longwave radiation and clouds enhance
the drying of the troposphere in regions 
of subsidence more than the cases when the longwave CRE is not active.  This is especially true for the 
simulations with a grid-spacing of 1 km and 2 km.  While the mid-tropospheric profiles of diabatic 
heating and cooling are similar between the LW CRE on/off simulations, 
there is a very strong response below about 850 hPa for the 25km model.  At that grid-spacing, with the 
LW CRE on, significant low-level clouds are formed, leading to a radiative cooling of almost 10K/day, 
and a radiative cooling of around -7K/day.   However, when the LW CRE is off, there is no appreciable 
difference in the low-level diabatic profiles among the E25, E2, and E1 experiments.   It is also clearly 
seen that the LW CRE decreases the upper-level ice-condensate for E25, E2, and E1 experiments but 
dramatically increases the low-level liquid condensate for E25 and P25 (Figure \ref{fig:TotCond}).

@@

The results presented in this paper show us that (scientific point learned).    Interactions between the clouds and radiation
act to spatially concentrate the circulation, strengthen the overturning circulation, and dry out the mid-troposphere where 
subsidence dominates (although domain mean condensate is less when LWCRE is on.).

It is remarkable that the control simulations (LW CRE on) have a precipitation rate  that can vary by as much as 
20\% despite have have the same prescribed SST and incoming radiation.  The simulations also all have the same radiation,
turbulence, large-scale cloud and microphysics parameterizations and all use the same dynamical core.

We were struck by three particular changes due strictly do a 4-fold increase in domain width.  A dramatic widening of the precipitation distribution when the LW CRE is off, a shifting of low level clouds upward by more than 100 hPa, and 
a strong dependence of the quantity of precipitation produced by the large-scale scheme on the LW CRE.

The longwave cloud radiative effect (LWCRE) has been shown to play a dramatic role in the organization of the precipitation
field and the low-level clouds.  This is primarily accomplished by decreasing the amount of precipitation that is produced 
by the large-scale precipitation scheme.  Two major differences between the GCM-like experiments and the CRM-like 
experiments are 4-5 times more low-level clouds and a much dryer mid-troposphere in the GCM-like experiments compared 
to the CRM-like experiments.  The LWCRE also highlights the impact of an increased vertical moisture transport 
in our CRM-like models.

Several different configurations of a mock-Walker circulation have been used to analyze the response of clouds and the 
large-scale circulation to a simple Gaussian shaped prescribed SST pattern.  To better understand the role that clouds
and humidity play in determining and driving the circulation we have performed experiments with and without the radiative
effect of clouds, with and without the deep convective parameterization, across multiple domain sizes,  
and using a range of resolutions from 100 km to 1 km grid-spacing.  Our results imply that the convective 
parameterization and the longwave cloud radiative effect strongly interact with each and often lead to 
asymmetric results.

Below is a list summarizing several of the changes that occur when the LW CRE is off.  Turning the LW CRE interactions off tends to:

\begin{itemize}
  \item decrease the strength of the overturning circulation and the domain mean precipitation (P100,P25,E25)
  \item increase the midtropospheric RH by $5-15 \% $
  \item eliminate the difference of liquid condensate at 900 hPa between explicit and parameterized simulations.
  \item weaken the horizontally oriented low-level circulations.   
  \item spread out in geographic space the precipitation maxima
  \item homogenize the surface enthalpy flux over the warm pool, primarily through the latent heat flux
  \item decrease both the LTS and the EIS, over both the warm and cold regions
  \item decrease the lowest level $q, \theta_e,$ temperature $T$ and virtual temperature $T_v$.  
  \item eliminate large differences in the evaporation and lowest level temperature in the center of the domain between 
           the 1 and 2 \textit{km} simulations. 
  \item increase evaporation in the 1 and 2 \textit{km} cases, decrease it in the 25 \textit{km} case.
  \item amplify the larger values of ice condensate aloft for the explicit simulations.  
\end{itemize}

Relative to simulations with a grid-spacing of $25 \textit{km}$, the E2 and E1 experiments can be characterized as having: 
\begin{itemize}
  \item stronger overturning circulations (as measured by vertical velocity) which are more consistently centered over the maximum of SST.  
  \item higher relative humidity in the upwelling regions and aloft.   Between 600-800 hPa the explicit models can have a relative humidity 
  that is 40\% higher than in the parameterized, lower resolution simulations.  
  \item two to four times more ice condensate above 600 hPa but only about half as much liquid condensate below 700 hPa.
  \item less domain mean precipitation.  Values for 1 and 2 km simulations are in the 3.2-3.5 mm/d range, while those for the 25 km 
  simulations are 10-20\% higher (3.5-3.9 mm/d).  
  \item a weaker radiative cooling rate (about 2 K/d) and weaker heating rate (about 2 K/d).
  %\item much less vertical shear.  
\end{itemize}

There is a rich literature on tropical overturning circulations.  Much of it is focused on the Hadley circulation and while some
work has looked at the Walker circulation, it has received less attention.  Part of our motivation in using the framework of the 
Walker circulation comes from the fact that all of the major types of tropical cloud types occur regularly within this circulation
clouds and the circulation.  While this study has interpreted the experiments in the context of the Walker Circulation, our results 
are also relevant to the deep overturning circulations and meridional SST gradients that define the ITCZ and the Hadley 
Circulation.  In that context, our results are consistent with those of several recent studies (e.g. Harrop and Hartmann, 2016, 
Popp and Silvers, 2017, Dixit et al. 2018, Flaschner et al. 2018).  These studies, as well as the present one, all show that the 
LWCRE acts to constrain the deep convective region.  This results from an increased atmospheric energy uptake and 
strengthening of the overturning circulation where the deep convective clouds occur (Popp and Silvers, 2017).  Consistent with 
these papers, the present work also shows that the LWCRE has a strong influence on the low-level circulation.   When 
the LWCRE is turned off, the low-level circulations shift upward and are not as well organized (reference psi figures).
There is a corresponding change in the low-level cloud fields, longwave radiative cooling, and the domain mean precipitation.
For the experiments with a GCM-like configuration, the LWCRE strongly influences the precipitation from the large-scale
cloud scheme while leaving the precipitation from the convective parameterization scheme largely unchanged.  This leads 
to a much stronger response of the GCM-like experiments to the LWCRE, especially in the low-levels of the troposphere.  
Because we expect the fraction of precipitation that is due to the convective parameterization to be model dependent 
this could explain why GCMs have a large spread in their response to turning off the LWCRE (reference some 
of Aiko's work?).

We have used the tropical overturning circulation to compare the multi-scale interactions between circulation patterns, cloud 
systems, and interactive radiation across experiments with grid-spacing ranging from 100km to 1km.  The flexible modeling
system at GFDL has allowed us to use a single code base in a GCM-like configuration with physics parameterizations 
that are very close to the AM4.0/CM4.0 models as well as in a CRM-like configuration with explicit convection.  While 
there are significant differences between the CRM presented in this paper and more conventional CRMs (e.g. vertical 
grid spacing and threshold based `binary' cloud scheme), the prospect of so easily converting a GCM into something 
like a CRM allows for an enticing testbed for future model development.
These comparisons have highlighted some of the unexpected behaviors of a GCM-like configuration when used with 
highly idealized boundary conditions.  For example, the consistent asymmetry of the circulation and precipitation relative 
to the fixed SST pattern and the dominance of the large-scale precipitation over the convective precipitation.  
The comparisons have also illustrated some 
of the challenges that arise when dramatically increasing the resolution of a GCM.  For example, the lack of shallow 
clouds (both convective and stratocumulus) and the difficulty of comparing clouds in this CRM to other CRMs due to the 
prognostic large-scale cloud scheme used at GFDL.

The goal in developing and using idealized models is to capitalize on their simplicity and learn something about the 
parent system as a result.  Idealized studies focusing on particular parts of the Earth system have shown that the danger of 
such idealized models is that they will not actually be less complex or easier to understand than the Earth system.

The only difference between our simulations and radiative convective equilibrium is a gradient of SST at the lower boundary.
The addition of this small difference creates an undeniable link with the observed tropical atmosphere.  The idealized 
mock -Walker cell configuration serves as an important step between models of pure radiative convective equilibrium
and models which are applicable to a wider range of Earth like conditions.

Although the radiative cooling and condensational heating
 are in many ways similar...   The 1 and 2 \textit{km} experiments transport much more condensate aloft and maintain a 
 relative humidity over the warm pool region that is about 40\% higher than the experiments with parameterized convection.

%References: 
%
%Bretherton et al. 2006
%Jeevanjee et al. 2018
%Kuang and Wofsy
%Muller and Held, 2012
%Popp and Silvers, 2017
%Patrizio and Randall, 2019
%Randall et al., 1989
%Silvers et al. 2016
%Slingo and Slingo, 1988
%Stevens et al. 2012
%Zhao et al. 2018a
%Zhao et al. 2018b

% PDF for location of precip maximum of precipitation for experiments at both 25km and 100km resolutions 
% LW CRE on
%\begin{figure}
%  \centering
%      %\includegraphics[width=0.96\columnwidth]{walkerfigs/hist_maxP_4mods_lwcreon.eps}
%       \includegraphics[width=0.96\columnwidth]{walkerfigs/hist_4pan_25vs100_lwcre.eps}
%  \caption{Normalized probability distributions of the location of the maximum value of precipitation for experiments 
%  with the longwave CRE on.  All panels show cases with parameterized convection, the top panels are computed 
%  for the experiments having a domain width of 4000 km while the bottom panels are for experiments having a 
%  domain width of 16000 km.  The black curves show a Gaussian function with a standard deviation computed for 
%  each experiment and an expected location in the center of the domain where the SST is warmest. }
%\end{figure}

>>>>

In contrast to the P100, P25, and E25 experiments just discussed, turning the LW CRE off for 
the E1 and E2 experiments results in more precipitation.  This can be understood as follows.  
The primary method by which the longwave cloud radiative effect influences the atmosphere is by 
heating the atmosphere in the region between the clouds and the surface.  
Larger values of ice condensate and upper-level cloud fraction as seen in the E2 and E1 experiment 
(Figure \ref{fig:TotCond}) therefore imply a larger atmospheric heating due to the CRE relative to the 
E25, P25, and P100 experiments in which there are fewer clouds aloft (Figures \ref{fig:TotCond}, 
\ref{fig:TotCond_P25P100}, and \ref{fig:cf_tdtlw}).  When the warming effect of the upper level clouds 
in the E1 and E2 experiments is removed in the LWCRE off experiments the energy balance of 
the atmosphere must be maintained through an increase of latent heating and subsequent increase 
of precipitation as reflected in Table 2.

@@

Water vapor is an efficient emitter of longwave radiation and thus greatly facilitates the cooling of the troposphere.  
\citeA{Popp2017} showed that there is dramatically less condensate in the atmosphere and much less precipitation 
(at the equator) for LW CRE off experiments (see their figure 1a,b).  Our results here for the E25, P25, and P100 
are consistent with those of \citeA{Popp2017} because the low-level condensate is strongly decreased 
when the LWCRE is off.  
%This is not the case for E2 and E1.  Those experiments have little low-level condensate to begin with 
%%and thus do not experience 
%%large changes in low-level condensate when the LWCRE is off.  
%but they do have a large increase of upper
%level condensate for the LWCRE off experiments.   
%This is a large difference in the way that models with explicit convection 
%respond to the LWCRE compared to more traditional large-scale models with parameterized convection.

The upper panels of Figures \ref{fig:rh_psi_P25E2E1} and \ref{fig:rh_psi_P25E2E1_lwoff} show that the 
interactions between longwave radiation and clouds enhance
the drying of the troposphere in regions of subsidence more than the cases when the longwave CRE is not active.  
This is especially true for the simulations with a grid-spacing of 1 km and 2 km.  While the mid-tropospheric 
profiles of diabatic heating and cooling are similar between the LW CRE on/off simulations, there is a very strong 
response below about 850 hPa for the 25km and 100km models.  At that grid-spacing, with the LW CRE on, significant low-level 
clouds are formed, leading to a radiative cooling on the order of -10K/day.   
However, when the LW CRE is off, there is little appreciable difference in the low-level diabatic profiles among the 
P100, P25, E25, E2, and E1 experiments.   It is also clearly seen that the LW CRE decreases the upper-level ice-condensate 
for E25, E2, and E1 experiments but dramatically increases the low-level liquid condensate for E25 and 
P25 (Figure \ref{fig:TotCond}).

@@

%The results presented in this paper show us that (scientific point learned).    Interactions between the clouds and radiation
%act to spatially concentrate the circulation, strengthen the overturning circulation, and dry out the mid-troposphere where 
%subsidence dominates (although domain mean condensate is less when LWCRE is on.).

We have used the framework of the tropical overturning circulation to compare the multi-scale interactions between 
circulation patterns, cloud systems, and interactive radiation across experiments with grid-spacing ranging from 
1km to 100km.  The flexible modeling
system at GFDL has allowed us to use a single code base in a GCM-like configuration with physics parameterizations 
that are very close to the AM4.0/CM4.0 models as well as in a CRM-like configuration with explicit convection.  While 
there are significant differences between the CRM presented in this paper and more conventional CRMs (e.g. vertical 
grid spacing and threshold based `binary' cloud scheme), the prospect of so easily converting a GCM into something 
like a CRM provides an enticing testbed for future model development.
These comparisons have highlighted some of the unexpected behaviors of a GCM-like configuration when used with 
idealized boundary conditions.  For example, the consistent asymmetry of the circulation and precipitation relative 
to the fixed SST pattern and the dominance of the large-scale precipitation over the convective precipitation.  
The comparisons have also illustrated some of the challenges that arise when dramatically increasing the resolution 
of a GCM.  For example, the lack of shallow 
clouds (both convective and stratocumulus) and the difficulty of comparing clouds in this CRM to other CRMs due to the 
prognostic large-scale cloud scheme used at GFDL.

%Several different configurations of a mock-Walker circulation have been used to analyze the response of clouds and the 
%large-scale circulation to a simple Gaussian shaped prescribed SST pattern.

To better understand the role that clouds
and humidity play in driving and responding to the circulation we have performed experiments with and without the radiative
effect of clouds, with and without the deep convective parameterization, across multiple domain sizes and resolutions.  
Our results show that the convective 
parameterization and the longwave cloud radiative effect strongly interact with each other and often lead to 
asymmetric results.

It is remarkable that the control simulations (LWCRE on) have a precipitation rate  that can vary by as much as 
20\% despite have have the same prescribed SST and incoming radiation.  All simulations have the same radiation,
turbulence, large-scale cloud and microphysics parameterizations and all use the same dynamical core.

We were struck by three particular changes due strictly do a 4-fold increase in domain width.  A dramatic widening of the precipitation distribution when the LWCRE is off, a shifting of low level clouds upward by more than 100 hPa, and 
a strong dependence of the quantity of precipitation produced by the large-scale scheme on the LWCRE.   
The LWCRE also highlights the impact of an increased vertical moisture transport 
in our CRM-like models.

%The longwave cloud radiative effect (LWCRE) has been shown to play a dramatic role in the organization of the precipitation
%field and the low-level clouds.  
%%This is primarily accomplished by decreasing the amount of precipitation that is produced 
%%by the large-scale precipitation scheme.  
%Two major differences between the GCM-like experiments and the CRM-like 
%experiments are 4-5 times more low-level clouds and a much dryer mid-troposphere in the GCM-like experiments compared 
%to the CRM-like experiments.  The LWCRE also highlights the impact of an increased vertical moisture transport 
%in our CRM-like models.

Below is a list summarizing several of the changes that occur when the LWCRE is turned off.  Turning the LWCRE 
interactions off tends to:

\begin{itemize}
  \item decrease the strength of the overturning circulation and the domain mean precipitation (P100,P25,E25)
  \item increase the midtropospheric RH by $5-15 \% $
 % \item eliminate the difference of liquid condensate at 900 hPa between explicit and parameterized simulations.
  \item weaken the horizontally oriented low-level circulations.   
  \item spread out in geographic space the precipitation maxima
  \item homogenize the surface enthalpy flux over the warm pool, primarily through the latent heat flux
  \item decrease both the LTS and the EIS, over both the warm and cold regions
  \item decrease the lowest level $q, \theta_e,$ temperature $T$ and virtual temperature $T_v$.  
  \item eliminate large differences in the evaporation and lowest level temperature in the center of the domain between 
           the 1 and 2 \textit{km} simulations. 
  \item increase evaporation in the 1 and 2 \textit{km} cases, decrease it in the 25 \textit{km} case.
  \item amplify the larger values of ice condensate aloft for the explicit simulations.  
\end{itemize}

Relative to simulations with a grid-spacing of $25 \textit{km}$, the E2 and E1 experiments can be characterized as having: 
\begin{itemize}
  \item stronger overturning circulations (as measured by vertical velocity) which are more consistently centered 
  over the maximum of SST.  
  \item higher relative humidity in the upwelling regions and aloft.   Between 600-800 hPa the explicit models 
  can have a relative humidity 
  that is 40\% higher than in the parameterized, lower resolution simulations.  
  \item two to four times more ice condensate above 600 hPa but only about half as much liquid condensate below 700 hPa.
  \item less domain mean precipitation.  Values for 1 and 2 km simulations are in the 3.2-3.5 mm/d range, while those for the 25 km 
  simulations are 10-20\% higher (3.5-3.9 mm/d).  
  %\item a weaker radiative cooling rate (about 2 K/d) and weaker heating rate (about 2 K/d).
  %\item much less vertical shear.  
\end{itemize}

There is a rich literature on tropical overturning circulations.  Much of it is focused on the Hadley circulation and while some
work has looked at the Walker circulation, it has received less attention.  Part of our motivation in using the framework of the 
Walker circulation comes from the fact that all of the major types of tropical clouds occur regularly within the region 
identified with the Walker circulation.  While this study has interpreted the experiments in the context of the Walker 
Circulation, our results are also relevant to the deep overturning circulations and meridional SST gradients that define 
the ITCZ and the Hadley Circulation.  In that context, our results are consistent with those of several recent studies 
(e.g. Harrop and Hartmann, 2016, 
Popp and Silvers, 2017, Dixit et al. 2018, Flaschner et al. 2018).  These studies, as well as the present one, all show that the 
LWCRE acts to constrain the deep convective region.  This results from an increased atmospheric energy uptake and 
strengthening of the overturning circulation where the deep convective clouds occur (Popp and Silvers, 2017).  Consistent with 
these papers, the present work also shows that the LWCRE has a strong influence on the low-level circulation.   When 
the LWCRE is turned off, the low-level circulations shift upward and are not as well organized (see Figures \ref{fig:rh_psi_P25E2E1} and \ref{fig:rh_psi_P25E2E1_lwoff}).
There is a corresponding change in the low-level cloud fields, longwave radiative cooling, and the domain mean precipitation.
For the experiments with a GCM-like configuration, the LWCRE strongly influences the precipitation from the large-scale
cloud scheme while leaving the precipitation from the convective parameterization scheme largely unchanged.  This leads 
to a much stronger response of the GCM-like experiments to the LWCRE, especially in the low-levels of the troposphere.  
Because we expect the fraction of precipitation that is due to the convective parameterization to be model dependent 
this could explain why GCMs have a large spread in their response to turning off the LWCRE (Aiko's work?).

The only difference between our simulations and radiative convective equilibrium is a gradient of SST at the lower boundary.
The addition of this small difference creates a concrete link with the observed tropical atmosphere.  The idealized 
mock -Walker cell configuration serves as an important step between models of pure radiative convective equilibrium
and models which are applicable to a wider range of Earth like conditions.   The goal in developing and using idealized models is to capitalize on their simplicity and learn something about the 
parent system as a result.  Idealized studies focusing on particular aspects of the Earth system have shown that a 
real danger of idealized models is that they will not actually be easier to understand than the Earth system.

What are we to conclude from this array of results?  Our CRMs succeed (by construction) at representing a far more 
realistic range of turbulent motions and thus presumably do a better job of explicit mixing compared to the GCM-like 
configurations.  However, the increase in resolved scales comes at the cost of most of the low-level clouds (\textit{what is 
the observed cloud fraction in the tropical Pacific below 700 hPa?}).  The GCM-like simulations match the CRM-like 
simulations  in the general two celled structure of the overturning circulation and the approximate distribution of condensate.  
While the GCM-like simulations succeed at producing 20-30\% low-level cloud fraction they produce very little upper-level 
clouds and the distribution of the precipitation is erratic and dependent on interactions between the convective 
parameterization and the interactions between longwave radiation and clouds.  The solutions of the GCM-like simulations 
also have a large dependence on the size of the domain.  Larger domains have more precipitation and domain mean 
cloud profiles shift upward by more than 100 hPa.  The amount of precipitation, and the structure of the precipitation and clouds 
is heavily dependent on the interaction between clouds and the longwave radiation.  This dependence strongly mediates
the large-scale cloud parameterization, rather than the convective parameterization.  The longwave cloud radiative 
effect acts to concentrate the regions of precipitation and increases the domain mean value.

We have learned that for high resolution models with explicit convection the upper-level clouds dominate the impact
of interactions between clouds and radiation while for GCM-like simulations it is the low-level clouds that dominate
this impact.

%The domain mean precipitable water is higher (as is P) when the LWCRE is on.   What does this imply?  Precipitation 
%efficiency or mixing?  The total water is PW plus condensates right?  So the Precipitation efficiency would only speak
%to the condensed part I think.

%References: 
%

%Kuang and Wofsy

%Randall et al., 1989
%Slingo and Slingo, 1988
%Stevens et al. 2012
